---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# positively

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/r-causal/positively/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/r-causal/positively/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

positively provides diagnostics for positivity violations and extrapolation in causal inference. Causal effect estimates rest on the positivity assumption: every unit must have a non-zero probability of receiving each level of the exposure within every stratum of the covariates. Where that probability is zero, or close to it, you face one of two problems: 1) for an exposure-based model, like Inverse Probability Weighting, you get a biased and unstable estimate or 2) for an outcome-based model, like a regression model, an estimate comes from the model extrapolating rather than from observed comparisons.

## Installation

positively is not yet on CRAN. You can install the development version from [GitHub](https://github.com/r-causal/positively) with:

``` r
# install.packages("pak")
pak::pak("r-causal/positively")
```

## Example

`check_positivity()` detects the exposure type, or uses the one you declare in `exposure_type`, and runs the diagnostics appropriate to that type. The bundled `pos_violations` dataset has two positivity problems planted in it: a structural subgroup that is never exposed, and a practical near-violation where the fitted propensity approaches zero or one in the tails of `x1`.

```{r example}
library(positively)

check <- check_positivity(pos_violations, exposure, c(x1, x2, region))
check
```

`sniff_violations()` gathers what the run found into one table, across every diagnostic, with the number behind each row in the `threshold` column where there is one.

```{r sniff}
sniff_violations(check)
```

The positivity regression tree (`port`) reports each problem as a rule over the covariates. Pull that diagnostic out with `$` and read its full results with `tidy()`, one row per reported subgroup. The `low_support` column marks those whose exposure prevalence falls below `beta` or above `1 - beta` among subgroups at least `alpha` of the sample.

```{r port-low-support, message = FALSE}
library(dplyr)

check$port |>
  tidy() |>
  filter(low_support)
```

The three low-support subgroups match the planted ground truth. One rule combining `region` and `x2` has an exposure prevalence of zero, the structural violation, and two `x1` rules in the lower and upper tails have prevalence near zero and one, the practical near-violation.

## Learn more

- Getting started: `vignette("positively")` walks through `check_positivity()` from end to end.
- The [package documentation](https://r-causal.github.io/positively/) holds the full reference and articles.
