---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# positively

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

positively provides diagnostics for positivity violations and extrapolation in causal inference. Causal effect estimates rest on the positivity assumption: every unit must have a non-zero probability of receiving each level of the exposure within every stratum of the covariates. Where that probability is zero, or close to it, an estimate comes from the model extrapolating rather than from observed comparisons.

## Installation

positively is not yet on CRAN. You can install the development version from [GitHub](https://github.com/r-causal/positively) with:

``` r
# install.packages("pak")
pak::pak("r-causal/positively")
```

## Example

`check_positivity()` detects the exposure type, or uses the one you declare in `exposure_type`, and runs the diagnostics appropriate to that type. The bundled `pos_violations` dataset has two positivity problems planted in it: a structural subgroup that is never exposed, and a practical near-violation where the fitted propensity approaches zero or one in the tails of `x1`.

```{r example}
library(positively)

check <- check_positivity(pos_violations, exposure, c(x1, x2, region))
check
```

The positivity regression tree (`port`) reports each problem as a rule over the covariates. Its plot draws every reported subgroup as a bar of exposure prevalence, with dashed reference lines at `beta` and `1 - beta` and the flagged subgroups highlighted.

```{r port-plot, fig.alt = "Horizontal bar chart of exposure prevalence for each PoRT subgroup, with flagged subgroups highlighted near prevalence zero and one."}
library(ggplot2)

port <- check_port(pos_violations, exposure, c(x1, x2, region))
autoplot(port)
```

The three flagged subgroups match the planted ground truth. One rule combining `region` and `x2` has an exposure prevalence of zero, the structural violation, and two `x1` rules in the lower and upper tails have prevalence near zero and one, the practical near-violation. positively locates these regions without grading them; the interpretation is yours.

## Learn more

- Getting started: `vignette("positively")` walks through `check_positivity()` from end to end.
- The [package documentation](https://r-causal.github.io/positively/) holds the full reference and articles.
