---
title: "Positivity for continuous exposures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Positivity for continuous exposures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%"
)
```

For a binary exposure, a positivity problem has a familiar shape: the fitted
propensity approaches zero or one, and there are two tails to inspect. A
continuous exposure has no single propensity to threshold. Positivity becomes a
question about support: for each candidate dose and each stratum of the
covariates, is there observed data near that dose? A violation is a region of
the joint exposure-covariate space that holds little or no data, and that region
does not reduce to two tails. It can sit in the middle of the exposure range, it
can depend on the covariates, and it can be invisible in the marginal
distribution of the exposure. The three continuous-exposure diagnostics in
positively measure support in three different ways, and each is informative about
a different geometry of violation.

## Setup

```{r setup, message = FALSE}
library(positively)
library(dplyr)
```

positively includes `pos_violations_long`, a simulated wide-format longitudinal
dataset with one row per subject over three time points. Each exposure column is
continuous, taking hundreds of distinct values.

```{r data}
pos_violations_long
```

The planted violation is documented in `?pos_violations_long` and lives in the
time-2 exposure `a2`. That exposure is drawn on `[0, 2]` or `[4, 6]`, so it never
falls in the open interval `(2, 4)`. The time-1 and time-3 exposures place values
throughout that interval, so the gap is specific to `a2` rather than a feature of
the exposure process as a whole.

```{r gap}
range(pos_violations_long$a2)
any(pos_violations_long$a2 > 2 & pos_violations_long$a2 < 4)
```

The sections below run each diagnostic on `a2` with the covariates `l0` and `l1`.
The gap is a mid-support hole in the marginal distribution of `a2`, so it is a
useful case for showing which diagnostic detects which kind of violation.

## Effective data points: support at chosen exposure values

`check_edp()` computes the effective-data-points diagnostic of Ring and Schomaker
(2026). For each observation moved to a candidate intervention value, it sums a
product kernel over the sample to measure how much observed data surrounds that
intervened-on point. The result runs from zero, meaning no observed support, up to
the sample size.

For a continuous exposure the default candidate values are the deciles of the
observed exposure, which follow the observed mass and here all fall outside the
gap. To probe the suspected hole, supply candidate values that span the range,
including one inside the gap.

```{r edp}
edp <- check_edp(pos_violations_long, a2, c(l0, l1), values = c(1, 3, 5))
edp
```

The message reports that the exposure was detected as continuous. Detection runs
once per call and announces the type it settles on, so a message of this form
accompanies every call below that leaves the type to detection, while a declared
type prints none. It is a default and not an override: the section on declaring
the exposure type shows how to set the type yourself when the heuristic reads a
continuous exposure as something else.

Effective data points are a relative measure, meaningful when compared across
candidate values for a fixed covariate set rather than against a universal
threshold. The candidate value `3` sits in the gap. Summarizing by candidate
value shows its effective data points are far below those at the supported values
`1` and `5`.

```{r edp-summary}
tidy(edp) |>
  group_by(value) |>
  summarize(edp = round(mean(edp), 1))
```

The `autoplot()` histogram, faceted by candidate value, shows the same result as
a distribution. The mass for the gap value `3` is concentrated at low effective
data points, while the supported values carry a broad distribution of well
supported observations.

```{r edp-plot, fig.alt = "Histograms of effective data points faceted by candidate exposure value one, three, and five. The value three panel concentrates at low effective data points, while values one and five spread toward higher support."}
ggplot2::autoplot(edp, type = "histogram")
```

Effective data points detect the `a2` gap directly, because the kernel measures
local support in the joint exposure-covariate space without assuming a form for
the conditional density.

## Hat values: leverage-based extrapolation risk

`check_hat_values()` computes the hat-value diagnostic of Moodie and Schulz
(2025). It fits the linear design matrix formed from the exposure and the
covariates, then for each exposure percentile and each observed covariate row it
measures the leverage that the resulting candidate point would exert on the model.
A candidate is flagged as high leverage when its hat value exceeds `threshold *
p / n`. The summary statistic is the proportion of high-leverage candidates,
calibrated against a null in which the exposure is drawn independently of the
covariates.

The null distribution is drawn by resampling, so we set a seed in the chunk. We
use `null_reps = 100` to keep the vignette fast. This is a toy setting; in
practice keep at least the default of 500 replicates so that the null quantile
is stable.

```{r hat-values}
set.seed(1)
hat_values <- check_hat_values(
  pos_violations_long,
  a2,
  c(l0, l1),
  null_reps = 100
)
hat_values
```

The observed summary statistic does not exceed the null quantile, so the leverage
profile is not distinguishable from one in which the exposure and covariates are
independent. Two properties of the diagnostic explain the result. First, the
candidate exposure values are percentiles of the observed exposure, which fall on
the two supported modes and never enter the gap. Second, `a2` does not depend
strongly on `l0` or `l1`, so no exposure-covariate combination is far from the
observed design in the leverage sense.

```{r hat-null, fig.alt = "Histogram of the null distribution of the leverage summary statistic, with a dashed line at the null quantile and a solid line at the observed statistic; the observed statistic sits within the null distribution."}
ggplot2::autoplot(hat_values, type = "null")
```

Hat values identify extrapolation risk that arises when specific exposure values
combine with covariate values in a way that is far from the observed design. That
risk is present when the exposure is covariate-driven, and it is evaluated at
plausible exposure percentiles rather than at holes in the exposure's own support.
The `a2` gap is a marginal mid-support hole, so it falls outside what hat values
measure.

## Highest density region non-overlap

`check_hdr()` computes the highest-density-region non-overlap ratio of Bao and
Schomaker (2025). For a common target exposure value, it reports the fraction of
covariate profiles whose highest-density region excludes that value, that is, the
share of the population for which the target dose is not supported. The ratio is
zero when the target is supported everywhere and one when it is supported nowhere.

```{r hdr}
hdr <- check_hdr(pos_violations_long, a2, c(l0, l1))
hdr
```

The non-overlap ratio is zero at every target, including targets inside the gap.
The default estimator, `hdr_density_normal()`, fits `lm(a2 ~ l0 + l1)` and treats
the conditional density as Gaussian around the fitted mean. The `a2` gap is a hole
between two modes of the conditional density, and the fitted normal fills that
hole and reports the target as supported. As documented in `?check_hdr`, the
default estimator detects mean-shift support gaps, where a stratum's supported
dose moves away from a target, and does not detect multimodal gaps. A flexible
conditional-density estimator supplied through `new_hdr_density()` is needed to
detect a multimodal hole.

To show the ratio when it is informative, consider a mean-shift example in which
the exposure depends strongly on a covariate.

```{r hdr-meanshift}
set.seed(2)
n <- 500
l <- rnorm(n)
dose <- rnorm(n, mean = 2 * l, sd = 0.4)
shift <- data.frame(dose = dose, l = l)

hdr_shift <- check_hdr(shift, dose, l)
hdr_shift
```

Here each covariate stratum supports a narrow band of doses around its own mean,
so no single common dose is supported across the whole population. The non-overlap
ratio has a nonzero floor even at the central target, and it rises toward one for
extreme targets. The `autoplot()` curve shows the ratio against the target value.

```{r hdr-plot, fig.alt = "Non-overlap ratio plotted against target exposure value for the mean-shift example, forming a U shape that is lowest near the center of the exposure range and rises toward one at both extremes."}
ggplot2::autoplot(hdr_shift)
```

The non-overlap ratio measures support for a common target dose across covariate
strata. It is a population-level quantity: a nonzero floor means that setting the
whole population to one dose is infeasible, not that the density model fits poorly.

## When each diagnostic is informative

The three diagnostics measure support in different ways, so they detect different
geometries of violation.

- Effective data points measure local support in the joint exposure-covariate
  space at candidate exposure values you choose. They detect sparse support at
  those values, including a mid-support hole in the exposure's own distribution,
  as with the `a2` gap.
- Hat values measure leverage-based extrapolation risk for exposure percentiles
  combined with observed covariate rows. They are informative when the exposure is
  covariate-driven and extreme dose-covariate combinations sit far from the
  observed design. They do not target holes in the exposure's marginal support.
- Highest density region non-overlap measures whether a common target dose falls
  within the highest-density region of each covariate stratum. With the default
  normal estimator it detects mean-shift gaps, where a stratum's supported dose
  moves away from the target, and requires a flexible density estimator to detect
  multimodal holes.

Running all three and reading them together shows both the location of a violation
and the kind of support it lacks.

## Declaring the exposure type

Every diagnostic infers the exposure type from the data unless you declare it.
Exactly two distinct values is read as binary; otherwise a numeric exposure is
read as categorical when fewer than 20 percent of its values are distinct, which
is a sensible default and the wrong answer for a coarsely measured dose.
Consider a drug dose recorded at eight milligram levels across 150 subjects,
with higher doses given to subjects with higher `l`.

```{r coarse}
set.seed(3)
n <- 150
l <- rnorm(n)
dose <- round(pmin(pmax(35 + 15 * l + rnorm(n, sd = 5), 0), 70) / 10) * 10
coarse <- data.frame(dose = dose, l = l)

sort(unique(coarse$dose))
```

The dose is continuous in every sense that matters to these diagnostics, but
eight distinct values over 150 observations fall well under the cutoff, so it is
detected as categorical.

```{r coarse-detected}
check_edp(coarse, dose, l, values = c(10, 40, 70))
```

`check_edp()` accepts a categorical exposure, so it runs regardless, treating
each dose level as a category. `check_hat_values()` and `check_hdr()` do not
accept one, so under detection alone neither would run here at all. Setting
`exposure_type` settles the question. A declared type is authoritative:
detection is not consulted at all, and the declaration is refused only when its
math cannot run on the column as given, which for `"continuous"` means a column
that is not numeric and for `"binary"` a column without exactly two distinct
values.

```{r coarse-declared}
check_hdr(coarse, dose, l, values = c(10, 40, 70), exposure_type = "continuous")
```

The non-overlap ratio is lowest at `40`, near the center of the observed doses,
and highest at the two extremes, which few covariate profiles support.

`check_positivity()` takes the same argument and forwards the resolved type to
every diagnostic it composes, so one declaration covers the whole set.

```{r coarse-composed}
set.seed(4)
check_positivity(coarse, dose, l, exposure_type = "continuous")
```

Each diagnostic offers only the types it supports, so `"binary"` is not among
`check_hdr()`'s choices and declaring it there fails on the argument rather than
on the column. `check_positivity()` takes all three types, so it is where the
structural rule does the refusing: eight distinct doses are not two.

```{r coarse-wrong, error = TRUE}
check_positivity(coarse, dose, l, exposure_type = "binary")
```

## Time-varying exposures

`check_hdr_seq()` applies the non-overlap ratio to a time-varying continuous
exposure, one time point at a time, fitting the conditional density of each
exposure on the history that enters its conditioning set. It accepts the same
wide-format data used above, with an ordered selection of exposure columns and a
per-time-point list of covariates. The *Identifying violations: PoRT and sPoRT*
vignette covers the sequential diagnostics in full, including conditioning-set
construction and the longitudinal support structure of `pos_violations_long`.

## Where to go next

The *Getting started with positively* vignette introduces `check_positivity()`,
the general entry point, and reads each diagnostic's output and plot on a binary
example. For a continuous exposure, `check_positivity()` runs effective data
points, positivity regression trees with the exposure categorized, hat values, and
highest density region non-overlap together, with the defaults used above.

## References

Bao Y, Schomaker M (2025). Feasible Dose-Response Curves for Continuous Treatments
Under Positivity Violations.

Moodie EEM, Schulz J (2025). A Simple Diagnostic for the Positivity Assumption for
Continuous Exposures. *Statistics in Medicine*, 44:e70194.

Ring C, Schomaker M (2026). A Diagnostic to Find and Help Combat Stochastic
Positivity Issues, with a Focus on Continuous Treatments.
