---
title: "Density ratios and modified treatment policies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Density ratios and modified treatment policies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%"
)
```

A modified treatment policy asks what would happen under an intervention that
shifts each subject's treatment as a function of their observed history, rather
than setting everyone to a fixed level. Estimators for these policies reweight
the observed data to the interventional world using a density ratio
$r = g^*(a \mid h) / g(a \mid h)$, the ratio of the intervention treatment law
to the observed treatment law. The ratio is large wherever the observed
conditional treatment density is small, which is where positivity is
threatened. Reading the distribution of these ratios is therefore a positivity
diagnostic for the policy.

`check_density_ratios()` summarizes user-supplied ratios. It accepts a numeric
vector for a point treatment, an n-by-T matrix whose columns are time points
for a time-varying treatment, and a fitted lmtp object whose
density-ratio component it reads. A `propensity::psw` weight vector is also
accepted.

## Setup

```{r setup}
library(positively)
```

## What the summaries measure

Because a density ratio is a Radon-Nikodym derivative, its mean is one whenever
the intervention law is absolutely continuous with respect to the observed law.
The mean is a specification check rather than a violation detector: well-behaved
weights on average do not guarantee positivity. The signal lives in the upper
tail and in the Kish effective sample size
$(\sum_i r_i)^2 / \sum_i r_i^2$, which equals the sample size when the ratios
are equal and falls as a few large weights come to dominate. Petersen and
colleagues (2012) recommend inspecting the weight distribution for exactly this
reason: a near violation of positivity appears as a heavy upper tail, while an
observed density close to zero for some treatment level produces ratios that
are large or exactly zero.

## A hand-built point-treatment example

To see the summaries against known behavior, construct two sets of ratios by
hand. The first is well-behaved: values scattered close to one. The second has
a heavy right tail. Both are rescaled to mean one so that the only difference
between them is the shape of the distribution. The seed is set so the example
is reproducible.

```{r hand-built}
set.seed(20260706)

well_behaved <- rlnorm(1000, meanlog = 0, sdlog = 0.3)
well_behaved <- well_behaved / mean(well_behaved)

heavy_tail <- rlnorm(1000, meanlog = 0, sdlog = 1.4)
heavy_tail <- heavy_tail / mean(heavy_tail)
```

Passing the well-behaved ratios prints a compact numeric summary: the number of
ratios and time points, the mean, the median and the upper quantiles, the
maximum, the proportion of ratios above each threshold, the proportion that are
exactly zero, and the Kish effective sample size as a fraction of the sample
size.

```{r well-behaved}
check_density_ratios(well_behaved)
```

The mean is one, as construction guarantees. The quantiles sit close together,
the maximum is a small multiple of the median, and the effective sample size
retains most of the 1000 observations. These weights carry no positivity
warning.

The heavy-tailed ratios have the same mean and the same sample size, but the
distribution is different.

```{r heavy-tail}
heavy_ratios <- check_density_ratios(heavy_tail)
heavy_ratios
```

The upper quantiles climb steeply, a small fraction of ratios exceed ten, and
the effective sample size fraction falls to roughly a fifth of the observations.
The estimate for this policy rests on a much smaller effective sample than the
raw count suggests, and it is sensitive to the handful of subjects carrying the
largest weights.

The `autoplot()` method draws the distribution of the raw ratios. The long
right tail is the visual counterpart of the collapsed effective sample size.

```{r heavy-tail-plot, fig.alt = "Histogram of hand-built density ratios with a long right tail extending well above the bulk of the distribution near one."}
autoplot(heavy_ratios, type = "distribution")
```

## A mean below one signals leaked support

A mean visibly below one is informative in its own right. It marks a structural
violation in which intervention probability mass falls outside the observed
support, so some ratios are exactly zero. The following ratios are rescaled to
mean one and then have a block of values set to zero, mimicking subjects for
whom the shifted treatment is never observed.

```{r degenerate}
set.seed(1)
degenerate <- rlnorm(500, meanlog = 0, sdlog = 0.5)
degenerate <- degenerate / mean(degenerate)
degenerate[sample(500, 40)] <- 0

check_density_ratios(degenerate)
```

The mean now prints below one. The proportion of exact-zero ratios is reported
as the `prop_zero` statistic, which the long summary from `tidy()` exposes
directly.

```{r degenerate-tidy}
tidy(check_density_ratios(degenerate)) |>
  dplyr::filter(statistic == "prop_zero")
```

## Time-varying treatments and cumulative products

For a time-varying treatment, supply one column per time point. The diagnostic
summarizes the per-time ratios and also their cumulative product across time
points, which is the density ratio of the whole treatment history. Mild
per-step non-overlap compounds multiplicatively, so the cumulative effective
sample size can fall even while each per-time fraction stays healthy.

```{r matrix}
set.seed(20260706)
per_time <- replicate(3, {
  ratios <- rlnorm(800, meanlog = 0, sdlog = 0.8)
  ratios / mean(ratios)
})

longitudinal <- check_density_ratios(per_time)
longitudinal
```

The printed per-time effective sample size fraction is moderate, but the
cumulative fraction is far lower: the three per-step distributions multiply into
a history weight with a much heavier tail. `glance()` returns the headline
cumulative statistics in one row.

```{r matrix-glance}
glance(longitudinal)
```

The `type = "cumulative"` plot shows the cumulative-product quantiles across
time points. The gap between the upper quantiles and the median widens with each
step, which is the sequential positivity signature.

```{r matrix-plot, fig.alt = "Cumulative-product quantiles of density ratios across three time points, with upper quantiles rising away from the median as time increases."}
autoplot(longitudinal, type = "cumulative")
```

## A fitted longitudinal policy with lmtp

In practice the ratios come from a fitted model. The lmtp package
estimates modified treatment policy effects and stores the estimated density
ratios in the fitted object. The chunk below is gated on lmtp being
installed, so the vignette builds either way.

The model fits a shift intervention that lowers each time-varying treatment by
one unit, on the `sim_t4` example data. The settings are chosen for a fast
demonstration: a subset of the rows, two cross-fitting folds, and a single
generalized linear learner. A serious analysis would use more folds and a
richer learner library.

```{r lmtp, eval = requireNamespace("lmtp", quietly = TRUE)}
library(lmtp)

set.seed(20260706)
demo_data <- head(sim_t4, 300)
treatments <- c("A_1", "A_2", "A_3", "A_4")
covariates <- list("L_1", "L_2", "L_3", "L_4")

shift_down <- function(data, trt) pmax(data[[trt]] - 1, 0)

fit <- lmtp_sdr(
  demo_data,
  trt = treatments,
  outcome = "Y",
  time_vary = covariates,
  shift = shift_down,
  outcome_type = "binomial",
  folds = 2,
  learners_trt = "SL.glm",
  learners_outcome = "SL.glm"
)

check_density_ratios(fit)
```

Passing the fit reads its density-ratio component and produces the same
summaries as a hand-built matrix. The per-time effective sample size fraction
describes the overlap at the final time point, while the cumulative fraction
describes the overlap for the full four-step history. A large gap between them
points to sequential positivity strain that no single time point reveals.

```{r lmtp-plot, eval = requireNamespace("lmtp", quietly = TRUE), fig.alt = "Cumulative-product quantiles of lmtp density ratios across four time points, showing the spread of the history weight widening over time."}
autoplot(check_density_ratios(fit), type = "cumulative")
```

## Inverse-probability weights from a propensity model

Inverse-probability weights for a binary treatment are also density ratios: they
reweight each treatment arm toward a target population. `check_density_ratios()`
accepts a `propensity::psw` vector, so weights from a fitted
propensity model can be inspected with the same summaries. The example uses the
`nhefs_weights` data from the halfmoon package, gated on that package being
installed.

```{r nhefs, eval = requireNamespace("halfmoon", quietly = TRUE)}
library(halfmoon)

check_density_ratios(nhefs_weights$w_ate)
```

Average treatment effect weights use a different normalization from a
mean-one modified treatment policy ratio, so their mean is near two rather than
one. The informative quantities are the same as before: the upper tail and the
Kish effective sample size fraction, which report how concentrated the weights
are and how much of the sample they leave effective.

## Where to go next

The *Getting started with positively* vignette introduces `check_positivity()`
and the diagnostics it composes for point treatments. `check_positivity()` does
not compose `check_density_ratios()`, because the ratios have to be supplied by
you rather than computed from the data, but `check_density_ratios()` reads and
plots through the same `tidy()`, `glance()`, and `autoplot()` interface as every
other diagnostic in the package.

## References

Petersen ML, Porter KE, Gruber S, Wang Y, van der Laan MJ (2012). Diagnosing and
responding to violations in the positivity assumption. *Statistical Methods in
Medical Research*, 21(1):31-54.
[doi:10.1177/0962280210386207](https://doi.org/10.1177/0962280210386207)

Ring C, Schomaker M (2026). A Diagnostic to Find and Help Combat Stochastic
Positivity Issues, with a Focus on Continuous Treatments.
