---
title: "Estimator-focused diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimator-focused diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%"
)
```

Most positivity diagnostics answer one question: does the assumption hold, and
where does support run thin? PoRT names the subgroups, effective data points
measure the support around each intervention, and the extrapolation check
locates units with no near neighbor in the opposite exposure group. The two
diagnostics in this vignette answer a different question. They take the
estimator you intend to use as given, and ask what a positivity violation does
to that estimator's answer. `check_eta_bias()` estimates the bias a chosen
estimator and truncation rule incur under the data's own fitted mechanisms.
`check_extrapolation()`, read through the lens of g-computation, shows which
units force the outcome model to predict outside the region where it was
trained.

## Setup

```{r setup}
library(positively)
```

Both examples use `pos_violations`, the simulated dataset described in
`?pos_violations` and in the getting-started vignette. It carries two planted
problems: a structural violation, where no subject with `region == "b"` and
`x2 > 1` is ever exposed, and a practical near-violation, where exposure depends
steeply on `x1`, so the fitted propensity approaches zero in the lower tail of
`x1` and one in the upper tail while both exposure levels remain observed.

## ETA.Bias: the bias an estimator incurs

`check_eta_bias()` implements the ETA.Bias parametric-bootstrap diagnostic of
Petersen et al. (2012). The idea is to treat the data's own fitted mechanisms as
a known truth and measure how a candidate estimator behaves under it.

The procedure has three parts. First, it fits a propensity model for the
exposure and an outcome regression on the exposure and covariates. The
g-computation contrast from the outcome model,
averaged over the sample, becomes the target `truth`. This target depends on the
outcome model alone, so it is the same for every estimator and constant across
truncation levels. Second, it draws bootstrap datasets: each resamples the
covariate rows, draws a bootstrap exposure from the untruncated fitted
propensity, and draws a bootstrap outcome from the fitted outcome model. Third,
it refits both models on each bootstrap dataset and applies the candidate
estimator, bounding the fitted propensity into the truncation interval where the
estimator uses it. ETA.Bias is the mean of the bootstrap estimates minus the
target. Because truncation is a property of the estimator rather than of the
data-generating mechanism, the bootstrap exposure is always drawn from the
untruncated propensity, so the diagnostic isolates the bias the estimator plus
its truncation rule introduce.

`check_eta_bias()` needs an outcome, which `pos_violations` does not include, so
`check_positivity()` never runs it and you call it directly. We construct a
continuous outcome that depends on the exposure and the covariates. Its true
average treatment effect is one.

```{r outcome}
set.seed(2024)
eta_data <- pos_violations
eta_data$y <- with(
  eta_data,
  exposure + x1 + 0.5 * x2 + rnorm(nrow(eta_data))
)
```

The covariates passed to `check_eta_bias()` must be numeric, so we supply `x1`
and `x2`. The steep dependence of the exposure on `x1` is the source of the
extreme propensities that inverse probability weighting reacts to. We diagnose
the inverse-probability-weighting estimator, which divides by the fitted
propensity and is therefore the estimator most exposed to a positivity
violation.

The truncation sweep is the useful mode. A vector of lower bounds passed to
`truncation_grid` becomes a set of truncation levels `c(lower, 1 - lower)`, and
every level reuses one shared set of bootstrap draws and model refits, so
differences across the sweep reflect the truncation bound alone. A lower bound
of zero applies no truncation.

The bootstrap is random, so we set a seed in the chunk. We use `n_boot = 25` to
keep the vignette fast. This is a toy setting; in practice use several hundred
bootstrap draws so that the Monte Carlo standard error is small relative to the
bias you are trying to read.

```{r eta-bias}
set.seed(2024)
eta <- check_eta_bias(
  eta_data,
  exposure,
  y,
  c(x1, x2),
  estimator = "ipw",
  truncation_grid = c(0, 0.01, 0.05, 0.1),
  n_boot = 25
)
eta
```

The printed output reports the target truth, the number of truncation levels,
and the range of ETA.Bias across the sweep. The full per-level table is in the
`@results` tibble, one row per truncation level with the bias, its Monte Carlo
standard error, and the mean bootstrap estimate.

```{r eta-results}
eta@results
```

Two patterns appear as the lower bound rises. The bias grows, because tighter
truncation replaces the extreme fitted propensities that carry the tail
observations with the truncation bound, which pulls the weighted estimate away
from the target. The Monte Carlo standard error shrinks, because bounding the
propensity away from zero and one stabilizes the weights. This is the
bias-variance tradeoff of weight truncation, read off a single fitted mechanism.

The `type = "sweep"` plot shows ETA.Bias against the truncation lower bound with
a band of plus or minus two Monte Carlo standard errors. The dashed line at zero
marks the bias a positivity-robust estimator would achieve.

```{r eta-sweep, fig.alt = "ETA.Bias plotted against the truncation lower bound for the inverse-probability-weighting estimator, rising away from zero as the bound tightens, with a two Monte Carlo standard error band."}
ggplot2::autoplot(eta, type = "sweep")
```

The three estimators respond differently. G-computation ignores the propensity,
so its ETA.Bias is Monte Carlo noise around zero and flat across the sweep. The
augmented, doubly robust estimator stays near zero in bias while carrying more
variance than g-computation. Inverse probability weighting is the estimator
whose bias the truncation sweep traces out, which is why we diagnose it here.
The `type = "bootstrap"` plot shows the distribution of bootstrap estimates at
each truncation level with the target truth marked, the raw material behind each
point on the sweep.

```{r eta-bootstrap, fig.alt = "Histograms of bootstrap estimates for the inverse-probability-weighting estimator, one facet per truncation level, each with a dashed line at the target truth."}
ggplot2::autoplot(eta, type = "bootstrap")
```

ETA.Bias captures the positivity, truncation, and sparsity component of bias and
excludes model-misspecification bias by construction. It reports a diagnostic
for a fitted mechanism and does not correct the estimate.

## Extrapolation as the g-computation view of positivity

G-computation predicts the outcome under each exposure value for every unit and
averages the contrast. It never divides by the propensity, so it does not
produce the extreme weights that a positivity violation creates for inverse
probability weighting. The violation does not disappear, though. It changes
form. Where one exposure group has no member near a unit's covariate profile,
g-computation must predict that unit's outcome under the missing exposure by
extrapolating the model beyond the data it was trained on. Under g-computation
the positivity problem becomes an extrapolation problem, and
`check_extrapolation()` measures it directly.

`check_extrapolation()` implements the extrapolation diagnostics of King and Zeng
(2006). For every observation it measures how well the opposite exposure group
covers that unit's position in covariate space, using Gower distances that
handle numeric and categorical covariates together. We pass all three
covariates, including the factor `region`.

```{r extrapolation}
extrapolation <- check_extrapolation(
  pos_violations,
  exposure,
  c(x1, x2, region)
)
extrapolation
```

The Gower distance between two rows is the mean over covariates of a
per-covariate distance: for a numeric covariate, the absolute difference divided
by the pooled sample range; for a factor, one when the categories differ and
zero when they agree. Distances lie in `[0, 1]`. The geometric variability is
one half the mean of the full distance matrix and sets the reference radius: an
observation counts an opposite-group unit as nearby when their Gower distance is
within one geometric variability, controlled by the `nearby` multiplier.

Three per-unit statistics are reported: `gower_min`, the distance to the nearest
opposite-group unit; `gower_mean`, the mean distance to the opposite group; and
`frac_nearby`, the fraction of the opposite group within the reference radius.

```{r extrapolation-results}
tidy(extrapolation)
```

The units with the lowest `frac_nearby` sit in the covariate tails and in the
structural subgroup, the same regions PoRT flags. The `type = "distribution"`
plot shows the distribution of `frac_nearby` within each exposure group. The
mass near zero is the set of units with no opposite-group neighbor close by, the
units whose g-computation prediction under the missing exposure rests on
extrapolation.

```{r extrapolation-distribution, fig.alt = "Histograms of the fraction of nearby opposite-group units, faceted by exposure group, each with mass near zero."}
ggplot2::autoplot(extrapolation, type = "distribution")
```

The convex-hull test is the sharper statement of the same idea. It asks whether
each observation lies inside the convex hull of the opposite group, solved as a
linear-programming feasibility problem on the numeric covariates. A unit outside
the opposite group's hull cannot be written as a weighted average of
opposite-group units, so any g-computation prediction for it under the missing
exposure is extrapolation by construction. The test runs automatically when
there are at most ten numeric covariates and the lpSolve package is
installed; lpSolve is a suggested dependency, so install it to enable the
hull test. The per-group summary reports the fraction of each group inside the
opposite group's hull.

```{r extrapolation-summary}
extrapolation@summary
```

The `type = "hull"` plot shows how many units of each group fall inside the
opposite group's hull. The units outside it are the ones that require
extrapolation, and they include the structural subgroup that is never exposed.

```{r extrapolation-hull, fig.alt = "Stacked bars per exposure group showing the share of units inside versus outside the opposite group's convex hull."}
ggplot2::autoplot(extrapolation, type = "hull")
```

Read together, the two diagnostics describe the same positivity violation from
the estimator's side. ETA.Bias shows the inverse-probability-weighting estimator
trading bias for variance across the truncation sweep, and the extrapolation
check shows the g-computation estimator leaning on model predictions for the
units in the tails and the structural subgroup.

## Where to go next

The getting-started vignette, *Getting started with positively*, covers
`check_positivity()` end to end and the diagnostics that answer whether
positivity holds. The vignette *Density ratios and modified treatment policies*
covers `check_density_ratios()`, the diagnostic for stochastic and modified
treatment policies, with hand-built ratios and with a fitted longitudinal model.

## References

King G, Zeng L (2006). The Dangers of Extreme Counterfactuals. *Political
Analysis*, 14(2):131-159.
[doi:10.1093/pan/mpj004](https://doi.org/10.1093/pan/mpj004)

Petersen ML, Porter KE, Gruber S, Wang Y, van der Laan MJ (2012). Diagnosing and
responding to violations in the positivity assumption. *Statistical Methods in
Medical Research*, 21(1):31-54.
[doi:10.1177/0962280210386207](https://doi.org/10.1177/0962280210386207)
