---
title: "Identifying Violations: PoRT and sPoRT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Identifying Violations: PoRT and sPoRT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%"
)
```

Most positivity diagnostics report a distribution of per-unit numbers and leave
you to decide which units matter. The Positivity Regression Tree (PoRT)
algorithm of Danelian et al. (2023) reports something different: named covariate
subgroups in which one exposure level is rare or absent, each described by a rule
you can read. A subgroup you can name is a subgroup you can reason about. You can
decide whether the rule describes a structural impossibility, which narrows the
causal question, or a practical shortage in the tails, which is a finite-sample
problem. This vignette covers `check_port()` for point treatments, the three
parameters that control its reading rule, continuous exposures, and the
sequential extension `check_port_seq()` for treatment regimes over time.

## Setup

```{r setup, message = FALSE}
library(positively)
library(dplyr)
```

We use the two simulated datasets that positively provides. `pos_violations`
is a point-treatment dataset with two planted problems, documented in
`?pos_violations`: a structural violation in which no subject with `region == "b"`
and `x2 > 1` is exposed, and a practical near-violation in the tails of `x1`,
where exposure is almost determined by `x1` while both exposure levels remain
observed overall. `pos_violations_long` is a wide longitudinal dataset with
continuous exposures `a1`, `a2`, and `a3` over three time points.

## Finding subgroups with PoRT

`check_port()` takes the data, the exposure column, and the covariates. It grows
shallow regression trees on the covariates with the exposure as the response,
then reads a fixed rule off each tree to report subgroups where exposure
prevalence is extreme.

```{r port}
port <- check_port(pos_violations, exposure, c(x1, x2, region))
port
```

The message reports that the exposure was detected as binary. For a binary
exposure the response is the indicator of the higher level, so `exposure_level`
is `"1"` and `prevalence` is the share exposed within each subgroup. PoRT reports
every terminal subgroup it examined, most of them well supported. The
`low_support` column marks the subgroups that meet the published reading rule: a
prevalence below `beta` (the prevalence threshold considered low) or above
`1 - beta`, among subgroups that are at least `alpha` of the sample. Filtering to
the low-support rows is the usual first step.

```{r port-low-support}
tidy(port) |>
  filter(low_support) |>
  select(description, exposure_level, n, prevalence)
```

PoRT recovers both planted violations and describes each in the covariates. The
rule `x2 >= ... & region = b` has an exposure prevalence of exactly zero: this is
the structural subgroup, empty by construction. The two `x1` rules, one in the
lower tail with prevalence near 0.05 and one in the upper tail with prevalence
near 0.96, are the practical near-violation, where exposure is almost but not
entirely determined by `x1`. The prevalence of exactly zero marks the structural
subgroup, and the near-zero and near-one prevalences mark the practical tails.

The `autoplot()` method draws the reported subgroups as a horizontal bar chart of
prevalence, with dashed reference lines at `beta` and `1 - beta` and the
low-support subgroups highlighted. Bar width is proportional to subgroup size. By
default the chart draws every reported subgroup, a couple of hundred here, which
crowds the y axis. Setting `low_support_only = TRUE` restricts the chart to the
low-support rows.

```{r port-plot, fig.alt = "Horizontal bar chart of exposure prevalence for the three low-support PoRT subgroups, drawn against dashed reference lines near prevalence zero and one."}
ggplot2::autoplot(port, low_support_only = TRUE)
```

## Tuning the reading rule

Three parameters govern which subgroups read as low support and how the trees are
searched. Reasoning about them is the core of using PoRT, because the same data
report different subgroups under different settings.

**`alpha`** is the minimum subgroup size for the reading rule, as a fraction of
the sample. Raising it holds PoRT to larger subgroups and suppresses small ones.
At the default `0.05` the rule flags subgroups of at least 50 observations; at
`0.10` it requires at least 100, which drops the upper `x1` tail and the
structural subgroup and leaves only the large lower tail.

```{r alpha}
port_alpha <- check_port(pos_violations, exposure, c(x1, x2, region), alpha = 0.10)
tidy(port_alpha) |>
  filter(low_support) |>
  select(description, n, prevalence)
```

**`beta`** is the prevalence threshold. A smaller `beta` demands a more extreme
prevalence before a subgroup is flagged. Setting `beta = "gruber"` resolves the
threshold to the sample-size-adaptive bound $5 / (\sqrt{n} \log n)$ that
sPoRT uses, which is about 0.023 at this sample size and so tighter than the
default. The tighter threshold drops the upper `x1` tail, whose prevalence of
0.96 no longer clears `1 - beta`, and tightens the lower-tail rule to a more
extreme cut of `x1`.

```{r beta}
port_beta <- check_port(pos_violations, exposure, c(x1, x2, region), beta = "gruber")
tidy(port_beta) |>
  filter(low_support) |>
  select(description, n, prevalence)
```

**`gamma`** is the maximum number of covariates that may jointly define a
subgroup. The combination search grows single-covariate trees first, then trees
over pairs of covariates, up to `gamma`. Raising it uncovers joint violations
that no single covariate isolates. The structural subgroup here is one such
violation: it is defined by `x2` and `region` together, so it appears only when
`gamma` is at least 2. At `gamma = 1` PoRT reports the two `x1` tails but not the
structural cell.

```{r gamma}
port_gamma <- check_port(pos_violations, exposure, c(x1, x2, region), gamma = 1)
tidy(port_gamma) |>
  filter(low_support) |>
  select(description, n, prevalence)
```

Higher `gamma` finds more, at the cost of more trees, and the combination search
is where PoRT spends its time on wide covariate sets.

## Continuous exposures

For a continuous exposure `check_port()` first categorizes the exposure into
bins, then treats each bin as a level and searches for covariate subgroups in
which that bin is rare. The number of quantile bins is set by `n_bins`, and
explicit cut points can be supplied through `breaks`, which overrides `n_bins`.
The following runs PoRT on the time-1 exposure `a1` from the longitudinal
dataset, conditioning on the baseline covariate `l0`, with four quantile bins.

```{r continuous}
port_cont <- check_port(pos_violations_long, a1, c(l0), n_bins = 4)
port_cont
```

The `exposure_level` column now names the bin interval rather than a level of a
binary exposure, and the reported subgroups describe regions of `l0` in which a
given bin of `a1` is rare. This locates conditional shortages of exposure across
the covariate. It does not target a gap in the marginal distribution of the
exposure itself; for continuous support gaps and their per-value diagnostics,
see *Positivity for continuous exposures*.

## Sequential violations with sPoRT

A treatment regime raises a problem the point diagnostic cannot see. A violation
at a later time point can be masked by subjects who were already treated and
carry that treatment forward. The sequential PoRT (sPoRT) algorithm of Chatton
et al. (2025) applies the reading rule one time point at a time. Under
`strategy = "stratified"`, the default, it fits a tree at each time point among
the subjects still following the regime, so a violation among those subjects is
not diluted by the already-treated. The second strategy, a pooled person-time
fit, is reserved for a later release; `strategy = "pooled"` currently signals an
error.

`check_port_seq()` takes the data in wide form. `.exposures` is an ordered
selection of the exposure columns, one per time point, and `.covariates` is a
list of selections, one element per time point, of the time-varying covariates.

```{r sport}
sport <- check_port_seq(
  pos_violations_long,
  c(a1, a2, a3),
  list(c(l0), c(l1), c(l2))
)
sport
```

At each time point the diagnostic runs the PoRT search with the current exposure
as the response and a conditioning set built from the baseline covariates, the
time-varying covariates, and the earlier exposures. The results carry a leading
`time` column, and the low-support rows report the time point at which each
subgroup was found.

```{r sport-low-support}
tidy(sport) |>
  filter(low_support) |>
  select(time, description, exposure_level, n, prevalence) |>
  head()
```

For a binary monotone regime the follower restriction changes the risk set at
each time point: it holds the subjects not yet treated, and it shrinks over
time. When `beta = "gruber"` the prevalence threshold is resolved separately at
each time point from the size of that risk set, so smaller follower sets are held
to a looser bound. The exposures in this dataset are continuous, so every subject
stays in every risk set, but the per-time trees and conditioning sets are built
the same way. The `autoplot()` method facets by time point, and each facet shows
the thresholds applied at its own time point. With `low_support_only = TRUE` only
the low-support subgroups are drawn, and a facet without bars means nothing read
as low support at that time point.

```{r sport-plot, fig.height = 6, fig.alt = "Faceted horizontal bar charts of low-support PoRT subgroup prevalence, one facet per time point, against per-time dashed reference lines."}
ggplot2::autoplot(sport, low_support_only = TRUE)
```

### Limiting the history with lag

The `lag` argument sets how far back the covariate and exposure history reaches
into each conditioning set. The default `Inf` uses the full history. A finite
`lag` restricts each conditioning set to the most recent time points, which
produces smaller conditioning sets and can drop rules that depend on distant
history. Setting `lag = 1` here keeps each time point's own covariates plus the
covariates and exposure of the one preceding time point, so the time-3 trees
condition on `l2`, `l1`, and `a2` alone. The time-1 covariate `l0` and exposure
`a1` leave the time-3 conditioning set, so low-support rules that referred to
`a1` can no longer appear, and the trees, refit on the smaller pool, report a
different set of subgroups.

```{r lag}
sport_lag <- check_port_seq(
  pos_violations_long,
  c(a1, a2, a3),
  list(c(l0), c(l1), c(l2)),
  lag = 1
)
tidy(sport_lag) |>
  filter(low_support, time == 3) |>
  select(time, description, exposure_level, n, prevalence)
```

### Censoring

In a longitudinal study, subjects who are censored leave the risk set, and
remaining uncensored is itself a condition of the counterfactual question. The
`.censoring` argument takes an ordered selection of censoring indicators, one per
time point, each coded `1` when the subject is censored and `0` otherwise. When
supplied, `check_port_seq()` analyzes each indicator as its own binary-exposure
tree in addition to the exposure trees, using the same two-sided reading rule.
The rule flags a subgroup whose censoring prevalence is below `beta` or above
`1 - beta`. The upper tail flags a subgroup nearly certain to be censored, just
as the exposure trees flag a subgroup nearly certain to be treated. The lower
tail flags a subgroup nearly certain to remain uncensored, the never-censored
subgroups on which the counterfactual under no censoring rests (Chatton et al.
2025). Because censoring is light in most subgroups, the lower tail usually
accounts for most of the low-support censoring rows. Following Chatton et al. (2025),
the censoring tree at each time point runs on the subjects uncensored through the
previous time point, the never-censored risk set, and adds the current treatment
to its conditioning set. The treatment rule plays no role in the censoring
process, so this risk set is the full uncensored cohort rather than the treatment
followers. Supplying `.censoring` also restricts each exposure risk set to the
followers who are still uncensored, since a subject censored earlier has no
observed exposure to analyze.

The longitudinal dataset carries continuous exposures and no censoring columns,
so the following small seeded example constructs a monotone binary regime with
censoring planted to be near-certain at time 2 for subjects with a baseline
covariate `l0` above 1.

```{r censoring-data}
set.seed(2025)
n <- 2000
l0 <- rnorm(n)
a1 <- rbinom(n, 1, 0.3)
c1 <- rbinom(n, 1, 0.08)

# Monotone treatment: once treated, stays treated.
a2 <- a1
followers <- a1 == 0
a2[followers] <- rbinom(sum(followers), 1, 0.4)

# Time-2 censoring among the uncensored, near-certain where l0 > 1.
c2 <- integer(n)
uncensored <- c1 == 0
prob_censor <- ifelse(l0 > 1, 0.98, 0.1)
c2[uncensored] <- rbinom(sum(uncensored), 1, prob_censor[uncensored])

regime <- data.frame(l0 = l0, a1 = a1, c1 = c1, a2 = a2, c2 = c2)
```

Passing `.censoring` runs both families of trees. The results carry a `type`
column that separates censoring rows from exposure rows.

```{r censoring-run}
cens <- check_port_seq(
  regime,
  c(a1, a2),
  list(c(l0)),
  .censoring = c(c1, c2)
)
cens
```

Filtering to the low-support censoring rows surfaces the planted subgroup at the
upper tail: at time 2, censoring is near-certain where `l0` is above 1, with a
prevalence close to 1. The remaining low-support rows sit at the lower tail, with
prevalences below `beta`. These are the never-censored subgroups, in which almost
no one is censored, and the two-sided rule reports them alongside the
near-certain-to-be-censored subgroup. Reading the `prevalence` column tells the
two tails apart.

```{r censoring-low-support}
tidy(cens) |>
  filter(type == "censoring", low_support) |>
  select(time, description, exposure_level, n, prevalence)
```

## Where to go next

- *Getting started with positively* introduces `check_positivity()` and the full
  set of point-treatment diagnostics on `pos_violations`.
- *Positivity for continuous exposures* covers effective data points, hat values,
  and highest density region non-overlap, which are the diagnostics designed
  around continuous support gaps.

## References

Chatton A, Schomaker M, Luque-Fernandez MA, Platt RW, Schnitzer ME (2025). Is
checking for sequential positivity violations getting you down? Try sPoRT!

Danelian G, Foucher Y, Léger M, Le Borgne F, Chatton A (2023). Identification of
in-sample positivity violations using regression trees: the PoRT algorithm.
</content>
</invoke>
