---
title: "Getting started with positively"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with positively}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%"
)
```

positively provides diagnostics for positivity violations and extrapolation in causal inference. Causal effect estimates rest on the positivity assumption: every unit must have a non-zero probability of receiving each level of the exposure within every stratum of the covariates. Where that probability is zero, or close to it, you face of two problems: 1) for an exposure-based model, like Inverse Probability Weighting, you get a biased and unstable estimate or 2) for an outcome-based model, like a regression model, an estimate comes from the model extrapolating rather than from observed comparisons.

Positivity violations come in two kinds, and they call for different responses.
A structural violation is one where treatment is impossible for a subgroup, so
no amount of additional data would fill the gap. A practical violation is one
where treatment is rare in a region of covariate space, so both exposure levels
remain observed overall but the propensity approaches zero or one in the tails.
Structural violations usually mean the causal question has to be narrowed to the
population where the exposure is possible. Practical violations are a
finite-sample problem that shows up as unstable weights and inflated variance,
and they can sometimes be managed by trimming, truncation, or a change of
estimand. Distinguishing the two is the first task a positivity diagnostic
faces, so the methods in positively locate the covariate regions responsible for
a violation in addition to measuring its overall severity.

## Setup

```{r setup, message = FALSE}
library(positively)
library(dplyr)
```

positively includes a small simulated dataset, `pos_violations`, with two known
problems planted in it so that a diagnostic can be checked against ground truth.
It has a binary `exposure`, two numeric covariates `x1` and `x2`, and a
two-level factor `region`.

```{r data}
pos_violations
```

The two planted violations are documented in `?pos_violations`. The first is
structural: no subject with `region == "b"` and `x2 > 1` is ever exposed,
because the true propensity is zero there. The second is practical: exposure
depends steeply on `x1`, so the fitted propensity drops below 0.01 in the lower
tail of `x1` and rises above 0.99 in the upper tail, while both exposure levels
stay observed across the sample as a whole. A good set of diagnostics should
recover both, and distinguish the empty subgroup from the near-empty tails.

## Running check_positivity()

`check_positivity()` is the general entry point. It detects the exposure type,
or uses the one you declare in `exposure_type`, then checks positivity with the
set of diagnostics appropriate to that type, composing the individual
`check_*()` functions with their defaults. It collects each diagnostic's output
into a container you can read and plot. You give it the data, the exposure
column, and the covariates. The covariates are required, with no default, so
that outcome columns are never swept in by accident.
`vignette("continuous-exposures")` covers when declaring the type is the right
call.

```{r check}
check <- check_positivity(pos_violations, exposure, c(x1, x2, region))
```

The message reports that the exposure was detected as binary. For a binary
exposure the default set is three diagnostics: effective data points (`edp`),
positivity regression trees (`port`), and an extrapolation check
(`extrapolation`). Printing the container shows each one in its own section.

```{r print}
check
```

Each section reports a numeric summary.
The `edp` section reports the range of effective data points across the sample.
The `port` section reports how many covariate subgroups were examined and how
many read as low support under the algorithm's published rule. The
`extrapolation` section reports the geometric scale used to judge nearness, the
average share of the opposite exposure group lying nearby each unit, and how many
units fall inside the opposite group's convex hull. The three low-support
subgroups and the units that sit far from any opposite-group neighbor are the
first signals that this dataset has positivity problems.

## Reading the container

`summary()` summarizes the container as a whole. It reports one row per
statistic per diagnostic: the `value` the diagnostic computed and, for the
statistics it pairs with a cut, the `threshold` behind that value. A threshold
is stated in the units of the quantity it cuts, which are not always the units
of the value beside it. The `port` count of low-support subgroups sits beside
`beta`, a prevalence, because the cut was applied to each subgroup's prevalence
in turn and the count is what came back. An `NA` means the overview has no one
number to show, not that the diagnostic read nothing against anything.

```{r summary-container}
summary(check)
```

To pull a single diagnostic out of the container, name it with `$` or `[[`. Its
`tidy()` method returns that diagnostic's own tidy tibble, one row per unit or
per subgroup depending on the method, and its `glance()` method returns the wide
row of statistics the overview drew on.

```{r tidy-child}
tidy(check$port)
```

Each diagnostic can also be called directly. Calling
`check_port()`, `check_edp()`, or `check_extrapolation()` directly returns the
same diagnostic object that `check_positivity()` composed, and each object
carries its own plot method through `autoplot()`. The sections below call the
individual functions so that we can plot and interpret each diagnostic in turn.

## PoRT: locating the subgroups

The positivity regression tree of Danelian et al. (2023) grows a shallow tree to
find subgroups where one exposure level is rare or absent. It is the most direct
diagnostic for the structural-versus-practical distinction, because it reports
each problem as a human-readable rule over the covariates.

```{r port}
port <- check_port(pos_violations, exposure, c(x1, x2, region))
port
```

The `low_support` column marks subgroups meeting the algorithm's published rule:
a prevalence below `beta` or above `1 - beta` among subgroups at least `alpha` of
the sample. Filtering to the low-support rows shows what PoRT found.

```{r port-low-support}
tidy(port) |>
  filter(low_support) |>
  select(description, n, prevalence, low_support)
```

PoRT recovers both planted violations and describes them in the covariates. The
rule `x2 >= ... & region = b` has an exposure prevalence of exactly zero: this
is the structural subgroup, the cell that is empty by construction. The two `x1`
rules, one in the lower tail with prevalence near 0.05 and one in the upper tail
with prevalence near 0.96, are the practical near-violation, where exposure is
almost but not entirely determined by `x1`. The `autoplot()` method draws the
reported subgroups as a bar chart of prevalence, with dashed reference lines at
`beta` and `1 - beta` and the low-support subgroups highlighted. The default
draws every reported subgroup; `low_support_only = TRUE` restricts the chart to
the low-support subgroups.

```{r port-plot, fig.alt = "Horizontal bar chart of exposure prevalence for the low-support PoRT subgroups, drawn near prevalence zero and one against dashed reference lines."}
ggplot2::autoplot(port, low_support_only = TRUE)
```

## Effective data points: how much support surrounds each intervention

The effective-data-points diagnostic of Ring and Schomaker (2026) measures, for
each unit moved to a candidate exposure value, how much of the observed sample
lies nearby in covariate space. It sums a product kernel over the data, so the
result runs from zero, meaning no observed support, up to the sample size. For a
binary exposure the candidate values are the two levels, 0 and 1.

```{r edp}
edp <- check_edp(pos_violations, exposure, c(x1, x2, region))
tidy(edp)
```

Effective data points are a relative measure. They are meaningful when compared
across units or across candidate values for a fixed covariate set, not against a
universal threshold. Units in a thinly supported region carry a low value at the
exposure level they rarely receive, because few observed neighbors surround the
intervened-on point. The `autoplot()` method shows the distribution of effective
data points, faceted by candidate exposure value; the left tail near zero is the
set of units with little observed support under that intervention.

```{r edp-plot, fig.alt = "Histograms of effective data points faceted by candidate exposure value, each with a left tail of low-support observations near zero."}
ggplot2::autoplot(edp, type = "histogram")
```

The planted regions make the comparison concrete. The following labels each
unit by the planted problem it belongs to: the structural subgroup is
`region == "b"` with `x2 > 1`, and the practical tails are the outer five
percent of `x1`. The labels are keyed by `.id`, the row identifier carried by
the per-unit results of `check_edp()` and `check_extrapolation()`, so they can
be joined to both sets of results below.

```{r planted-regions}
planted <- pos_violations |>
  mutate(
    .id = row_number(),
    planted = case_when(
      region == "b" & x2 > 1 ~ "structural",
      x1 < quantile(x1, 0.05) | x1 > quantile(x1, 0.95) ~ "x1 tail",
      .default = "elsewhere"
    )
  ) |>
  select(.id, planted)

planted |>
  count(planted)
```

Comparing effective data points at the exposed level across those labels shows
the thin support directly.

```{r edp-planted}
tidy(edp) |>
  filter(value == 1) |>
  left_join(planted, by = ".id") |>
  group_by(planted) |>
  summarize(edp = median(edp))
```

The median unit elsewhere in the sample has about 73 effective data points
under the exposed intervention. The planted regions carry a fraction of that:
about 21 in the structural subgroup and about 11 in the `x1` tails, the same
regions PoRT reported as rules.

## Extrapolation: distance and the convex hull

The extrapolation check follows the logic of King and Zeng (2006): a unit whose
covariate profile has no near neighbor in the opposite exposure group can only
be compared by extrapolating the model beyond the observed data. It measures
nearness with a Gower distance that handles numeric and categorical covariates
together, and, when the covariate count is small enough, tests whether each unit
falls inside the convex hull of the opposite group.

```{r extrapolation}
extrapolation <- check_extrapolation(pos_violations, exposure, c(x1, x2, region))
tidy(extrapolation)
```

`summary()` aggregates the results by exposure group. It reports, for each
group, the average distance to the nearest opposite-group unit, the share of
units with a neighbor within one geometric variability, and the share inside the
opposite group's hull.

```{r extrapolation-summary}
summary(extrapolation)
```

The same planted labels trace the extrapolation results to the unit level.

```{r extrapolation-planted}
tidy(extrapolation) |>
  left_join(planted, by = ".id") |>
  group_by(planted) |>
  summarize(
    frac_nearby = median(frac_nearby),
    in_hull = mean(in_hull)
  )
```

The `x1` tails are the extreme case on both measures. The median tail unit has
about a tenth of the opposite group nearby, against a third elsewhere, and
about a fifth of tail units fall inside the opposite group's hull, against 86
percent elsewhere. The structural subgroup sits between the two: its units
have fewer opposite-group neighbors than the rest of the sample, but the hull
test runs on the numeric covariates alone, so a subgroup defined partly by
`region` separates less sharply here than in PoRT, which searches the
categorical covariate directly. The `type = "distribution"` plot shows the
distribution of `frac_nearby` within each exposure group; the mass near zero is
the set of units with no opposite-group neighbor close by.

```{r extrapolation-distribution, fig.alt = "Histograms of the fraction of nearby opposite-group units, faceted by exposure group, each with a spike near zero."}
ggplot2::autoplot(extrapolation, type = "distribution")
```

When the convex-hull test runs, the `type = "hull"` plot shows how many units of
each group fall inside the opposite group's hull. Units outside the hull require
extrapolation by construction.

```{r extrapolation-hull, fig.alt = "Stacked bars per exposure group showing the share of units inside versus outside the opposite group's convex hull."}
ggplot2::autoplot(extrapolation, type = "hull")
```

The three diagnostics tell one story about the planted problems, each at its
own resolution. PoRT names the structural subgroup and the practical tails as
explicit rules. Effective data points shrink to a fraction of the sample
median in both regions under the exposed intervention. The extrapolation check
places the `x1` tails far from any opposite-group neighbor and largely outside
the hull, and it ranks the structural subgroup below the rest of the sample on
nearby support, with the caveat that its hull test reads only the numeric
covariates.

## Choosing diagnostics beyond the defaults

The default set depends on the exposure type. A categorical exposure drops the
extrapolation check; a continuous exposure adds hat values and highest density
regions and categorizes the exposure for PoRT. You can also request an explicit
subset with the `diagnostics` argument and tune any method through `args`, a
named list of per-method option lists.

```{r subset}
check_positivity(
  pos_violations,
  exposure,
  c(x1, x2, region),
  diagnostics = "port",
  args = list(port = list(alpha = 0.1))
)
```

Two diagnostics are never run by `check_positivity()`, because each needs
something the entry point does not have. `check_eta_bias()` needs an outcome,
and `check_density_ratios()` needs user-supplied density ratios. Call both
directly.

The other vignettes go deeper into each area:

- *Positivity for continuous exposures* covers effective data points, hat
  values, and highest density region non-overlap, and when each one applies.
- *Identifying Violations: PoRT and sPoRT* covers subgroup trees, tuning `alpha`,
  `beta`, and `gamma`, and the longitudinal and censoring variants.
- *Estimator-focused diagnostics* covers ETA.Bias with the truncation sweep and
  g-computation extrapolation.
- *Density ratios and modified treatment policies* covers
  `check_density_ratios()` with hand-built ratios and with a fitted longitudinal
  model.

## References

Danelian G, Foucher Y, Léger M, Le Borgne F, Chatton A (2023). Identification of
in-sample positivity violations using regression trees: the PoRT algorithm.

King G, Zeng L (2006). The Dangers of Extreme Counterfactuals. *Political
Analysis*, 14(2):131-159.

Ring C, Schomaker M (2026). A Diagnostic to Find and Help Combat Stochastic
Positivity Issues, with a Focus on Continuous Treatments.
</content>
